Adapting to Quasi-Periodic and Secular Variation of Climate Risk

### 2018-03-29 writeup/main.tex
Updates to text, primarily case study and intro

\title{Adapting to Quasi-Periodic and Secular Changes of Climate Risk}
\author[1,2]{James Doss-Gollin}
\author[1,2]{Upmanu Lall}
\author[1,2]{David Farnham}
\affil[1]{Columbia Water Center, Columbia University}
\affil[2]{Department of Earth and Environmental Engineering, Columbia University}

\input{header/biblatex.tex}
\addbibresource{library.bib}
\graphicspath{{../figs/}}

@@

\begin{document}
\maketitle
\RaggedRight{}
\begin{abstract}
  Hydroclimatic systems vary on many timescales, including regime-like variability and secular variability  \citep[``nonstationarity'';][]{Milly2008,Merz2014,Hodgkins2017}\footnote{Some people don't like citations in the abstract unless the paper is a rebuttal to another -- in which case citing the original is probably good. My impression is that our citations int he abstract are not highly necessary so I may take them out. Up to you.}.
  This multi-scale variability greatly complicates the estimation of future flood sequences, leading to ``low confidence in projections of changes in fluvial floods'' \citep{IPCC2012}.
  Despite broad recognition of nonstationarity's importance, there has been little theoretical development of ideas for the design and operation of infrastructure that simultaneously consider the potential for predictability of secular change and regime behavior or low-frequency variability.
  Consequently, existing theories of flood risk assessment are well-suited to analysis of flood adaptation strategies with fixed project operation period $M$, but not for analysis of sequential decision strategies.
  In this paper we illustrate some considerations of uncertainty and risk analysis associated with the sequential decisions that may need to be made for climate risk adaptation in a nonstationary setting.
  We simulate synthetic data and fit the resulting simulations in a Bayesian context using stationary and non-stationary models to show \(\ldots{}\).
  This is a first conceptual step to the development of a full sequential decision analysis approach under nonstationarity.
\end{abstract}

@@

\section{Introduction}

In 2011 four typhoons struck northern Thailand \citep{Gale2013}, causing severe damage to public and private infrastructure and crippling global supply chains \citep{Haraguchi2015}.
In June 2013, a rainy spring was followed by three summertime cyclones following highly similar tracks, leading to widespread flooding in central Europe \citep{Grams2014}.
At the same time a multi-year drought in S\~ao Paolo threatened to dry out one of the world's largest cities \citep{Seth2015}, and more recently the 2017 Atlantic Hurricane season led to approximately \usd{100 billion} in insured losses.\footnote{\url{http://www.swissre.com/media/news_releases/nr20171220_sigma_estimates.html}}
Since classical analysis of climate extremes for use in risk analysis and mitigation studies assumes that  the historical or simulated record of events constitutes an independent and identically distributed process, adaptation systems may be overwhelmed by events which violate these assumptions.\footnote{Do we need to add something about climate mechanisms such as wave resonance, blocking, \etc{} that can explain some of this? Or just cite \citet{Hannachi2017} to say the atmosphere has regimes?} \footnote{As for $\uparrow$, I would avoid the impulse to get too into the dynamics in this paper (i.e. wave resonance, etc). However, I do think that this paragraph could be a bit more direct. There is ample evidence that climate extremes are not IID $\rightarrow$ for example, Thailand, etc $\rightarrow$ thus risk analyses need to consider this... Basically my critique is that three sentences about recent disasters without the context of why we particularly care about these things may be a bit of a misdirection for the reader}.

Indeed, the scientific literature has recognized that even small nonstationarity (temporal change in dynamics of the underlying physical process) can lead to significant trends in the estimated $T$-year return periods or probability of exceedance $p_T$ \citep{Milly2008,Vogel2011,Salas2014,Merz2014,Obeysekera2016}.\footnote{All of these citations are for floods, can we look at other types of extremes? Even droughts would be helpful.}
This has led to a plethora of efforts aimed at identifying the nature of the nonstationarity and modeling it, \eg{} by introducing time trends on statistical parameters \citep{Strupczewski2001,Vogel2011,Obeysekera2014,Serinaldi2015}, climate covariates \citep{Delgado2014,Griffis2007,Silva2016,Sun2014,Hall2014}, latent variable or state-based modeling approaches \citep{Griffis2007,Sveinsson2005,Waylen1986}, or using only the last $N$ years of the data.\footnote{Rich Vogel citation?}
Each model will necessarily give different estimates, which may fit the historical record with comparable accuracy but diverge at an increasing rate as the model is forecast further into the future.
Despite the wide suite of available models, \emph{the question of how one should use these nonstationarity estimates for the design of risk mitigation infrastructure or the pricing of insurance instruments has been much less studied}.

There are two kinds of nonstationarities of concern.
The first, secular change, may be related to anthropogenic climate change, river modification, or land use change \citep{Merz2014}.
The uncertainties associated with this are not purely informational; both nature and future human actions will determine the outcomes, and neither is obviously predictable.
The approach taken by many, including the IPCC, is to consider discrete scenarios for what humans may do with respect to greenhouse gas emissions and then use a sequence of physics-based numerical models with these scenarios to develop probabilistic statements as to the potential changes in the extremes.
Such a ``model chain'' typically encompasses:
\begin{enumerate*}[label= (\roman*) ]
  \item emission scenario;
  \item general circulation model (GCM);
  \item downscaling; and
  \item high-resolution model of the study region.
\end{enumerate*}
Because of inherent errors due to, \eg{} the discretization and parameterization of the equations of motion, bias correction methods are typically applied at each step of the chain; while this typically improves the model chain's representation of the historical record, the extrapolation of these bias corrections far into the future does not inspire confidence\footnote{Is "inspire confidence" too informal?... I'm not sure}. \citep{Dankers2009,Ott2013,Merz2014,Dittes2017}.\footnote{Farnham et al 2018!}
A critical question in this regard is how to assess whether any structure\footnote{or other risk management instrument such as an insurance product?} designed using these projections may actually perform in the future relative to the proposed level of risk protection.
Would it be over- or under- designed?
Would the bias and uncertainty in its estimation lead to sub-optimal investment strategies in risk mitigation?
These questions go far beyond what can be reliably answered at this time, and this type of epistemic uncertainty clearly grows as we look out into the distant future.\footnote{Perhaps these sentences belong elsewhere?}

In addition to secular change, quasi-periodic inter-annual to multi-decadal modes of climate variability \citep[see][]{Hannachi2017} to influence the spatio-temporal expression of climate extremes in ways that violate the assumption of independent, identically distributed events \citep[see][]{Merz2014,Serinaldi2015,Hoskins2015}.
For example, even when no secular trend exists, using the past $N$ years to project the next $N$ leads to different estimates of floods relative to when the time series is fully shuffled to remove temporal clustering, even for very long ($N=\SI{200}{years}$) sequences \citep{Jain2001}.

When secular and quasi-periodic variability are both considered, it becomes clear the probability of exceedance is temporally variable over the life of a climate adaptation project.
In this paper we provide some initial insight into the bias and uncertainty associated with flood frequency estimation for projects as a function of the project planning period $M$ as well as the length of the historical record $N$.
This project planning period can be interpreted as the design life of a physical structure, or as the contract length of a financial instrument.
Our aim is not to identify the ``best'' model, but rather to develop theory as to how the underlying nature of the natural variability and nonstationarity and the choice of model used to model future flood risk map onto estimation bias and variance in the context of many different $M$ and $N$.
This estimation bias and variance may map directly onto the price of an insurance contract, or the possibility of over/under-design of a large structure like a dam or seawall used to protect against climate hazards.
Though the specific examples target flood risk, this method and framework may be used for understanding the implications of uncertain projections of the future for climate hazards including drought, hurricanes, and wind. \footnote{Overall comment on the introduction:  We may consider hinting at some of the results a bit more, otherwise the introduction is fairly wide open and I'm not sure the reader gets a good sense for what she is in for... but then again, perhaps that's OK}.

@@

To generate $\mu(t)$, a sequence of length $M+N$ is randomly chosen from the \SI{100000}{year} NINO3 sequence.
Next, $\mu(t)$ is calculated deterministically from the NINO3 time series $x(t)$ by
\begin{equation}
  \mu(t) = \mu_0 + \beta x(t) + \gamma \qty(t - t_0)
\end{equation}
where $\beta$ controls the sensitivity to the NINO3 variation and $\gamma$ controls the magnitude of the secular trend ($\gamma=0$ gives a stationary series).

@@

Analysis of the variance, in these estimates, shown in \cref{fig:stationary-variance} as the log standard deviation, is similar.
As suggested in the bias plots, the HMM and LN2 Stationary models have very low variance of their estimates.
The LN2 Trend model has much higher variance in estimates, particularly for short $N$ or short $M$.
\begin{figure}[ht]
  \includegraphics[width=\textwidth]{variance_stationary.pdf}
  \caption{The estimation variance for 1000 sequences generated by each of two stationary generating functions (row) and fit by each of three fitting functions (columns) as a function of $M$ and $N$.\label{fig:stationary-variance}}
\end{figure}

@@

\section{Case Study}

I would go back to a case study and look at a 500-year streamflow sample (could be generated synthetically if we can't get a good real one).
Fix $N=100$ (say) and choose between $M=5$ or $M=50$. 
Estimate this model with 
\begin{enumerate}[label=(\roman*)]
  \item HMM;
  \item LN2 Stationary;
  \item LN2 Trend; and
  \item Perfect Information (\ie{} knowing the future exactly)
\end{enumerate}
Then at each time, calculate the insurance price for each model from \cref{eq:insurance-price} (the perfect information model will have a risk premium of zero) and summarize the over or under-pricing of the risk premium.
Then explain that one would be unlikely to buy a 50-year insurance contract but that it's helpful to compare the two models using equal criteria.
We could extend to include NPV analysis if desired.

\emph{Please comment: should this be part of our paper?}

@@

Infrastructure in the developed world is aging \citep{Ho2017}, leading to questions of whether and how to replace it.
In the developed world new infrastructure is being built.\footnote{citations}
Given a positive future discount rate and uncertainty associated with future urbanization, economic growth, and climate change, the expected net benefits of this infrastructure are highly uncertain.
At the same time the private sector has accepted an increasing share of global risk \citep{WorldEconomicForum2016}, which it has mitigated primarily through financial instruments.
How should different contracts be priced? When does it make sense to build large infrastructure?
These questions cannot be answered with existing methodologies.

@@

>>>>

\title{Adapting to Quasi-Periodic and Secular Variation of Climate Risk}
\author[1,2]{James Doss-Gollin}
\author[1,2]{David Farnham}
\author[1,2]{Upmanu Lall}
\affil[1]{Columbia Water Center, Columbia University}
\affil[2]{Department of Earth and Environmental Engineering, Columbia University}

\input{header/glossaries.tex}
\input{header/biblatex.tex}
\addbibresource{library.bib}
\graphicspath{{../figs/}}

@@

\begin{document}
\maketitle
\RaggedRight{}

\begin{abstract}
  Recent disasters highlight the need for improved adaptation to extreme floods and other climate hazards.
  Climate risk mitigation has traditionally been achieved through large-scale infrastructure projects with long project lifespans.
  However, political and economic constraints limit the ability of central governments to fund such projects, leading to an increased role for the private sector and local governments \citep{WorldEconomicForum2016}.
  These actors typically operate on much shorter planning horizons for which the quasi-periodic variations of hydroclimate systems may be of interest.
  We use a series of structured experiments to explore how estimated and true climate risk vary as a function of the project planning period and the available data record.
  Strong differences in the expected value of flood risk, and the associated uncertainty, emerge between long planning periods dominated by secular trend and short planning periods dominated by quasi-periodic variability.
  Successful climate adaptation strategies should consider decisions on multiple timescales.
\end{abstract}

@@

\section{Introduction}\label{sec:introduction}

The \nth{20} century saw the evolution of a nearly universal approach to the statistical analysis of weather and climate hazard.
In this framework, event risk is assessed as a relationship between the event magnitude and its annual probability of exceedance.
Under the assumption that the historical or simulated record of events constitutes an stationary, \gls{iid} process, and using an appropriate statistical inference model, an estimate of this relationship and its uncertainty based on a finite $N$-year record is developed.
This is then used as a measure for decision analysis for infrastructure, flood zoning and insurance instruments for event risk mitigation applied to the costs and residual losses expected from the mitigation instrument.
The uncertainty associated with the model may be used to develop a risk premium for insurance, or for assessing the potential over- or under- design of infrastructure.

Over time it has become clear that even small fluctuations in the dynamics of the hydroclimate system, known as nonstationarities \citep{Milly2008}, can lead to clustering of extreme events  and strong time dependence of a public or private actor's risk profile \citep{Bonnafous2017a,Bonnafous2017b}.
For example, in 2011 four typhoons struck northern Thailand \citep{Gale2013}, causing severe damage to public and private infrastructure and crippling global supply chains \citep{Haraguchi2015}.
Similarly, in June 2013 three summertime cyclones followed highly similar tracks, leading to widespread flooding in central Europe \citep{Grams2014}.
More recently the 2017 Atlantic hurricane season led to over \usd{130 billion} in insured losses \citep{SwissRe2017}, despite negligible losses in the preceding several years.
Each of these events was highly improbable under \gls{iid} assumptions, yet the clustering of such events is well-documented in the literature \citep{Hurst1951,Matalas2012,Merz2014} and is consistent with our understanding of the nonlinear dynamics of the climate system \citep{Hannachi2017}.

Nonstationarity behavior that leads to violation of traditional assumptions takes two primary forms.
The first, secular change, may be related to anthropogenic climate change, river modification, or land use change \citep{Merz2014}.
The uncertainties associated with this are not purely informational, at least in the sense that they can be inferred from a historical sample or model.
Both nature and human actions over the course of this century will determine the magnitude and direction of secular change, and neither is fully predictable.
The uncertainty associated with secular change clearly grows as one looks out into the distant future, and the importance of investment decisions that may engage us in considerations of designs that go far out into the future, \eg{} 2100, diminish if any positive discount rate is used for the future.
While this consideration offsets the impact of the growing uncertainty for longer project life, it also suggests that it may be more effective to understand the consequences for analysis that may focus on shorter lived adaptation strategies which may last for 1, 5, 10, 20,  30 or 50 years.

Focusing on these shorter time horizons brings up the concern associated with the second type of nonstationarity in climate: the quasi-periodic inter-annual, decadal and multidecadal modes of climate variability that influence the spatiotemporal expression of climate extremes in ways that violate the standard \gls{iid} assumption \citep{Matalas2012,Hannachi2017}.
For example, analysis of a long synthetic data set demonstrated that using the past $N$ years to estimate return periods for the next $N$ years leads to a high degree of surprise even when secular change is exactly zero \citep{Jain2001}.
The finding that floods at a particular location are clustered in time relative to what could be expected under an \gls{iid} process was replicated for an observed long streamflow record \citep{Jain2001} and is consistent with analysis of a \SI{1500}{year} streamflow reconstructions \citep{Swierczynski2012}.
Analysis of annual-maximum rainfall events at sites around the world has also demonstrated a high degree of spatial clustering \citep{Bonnafous2017a}.

Broad recognition of nonstationarity has led to a plethora of efforts aimed at identifying the nature of the nonstationary and modeling it, though no single approach has proven broadly successful.
Such efforts have included time trends on statistical parameters \citep{Strupczewski2001,Vogel2011,Obeysekera2014,Serinaldi2015}, climate covariates \citep{Griffis2007,Sun2014,Hall2014,Delgado2014,Silva2016}, latent variable or state-based modeling approaches \citep{Waylen1986,Sveinsson2005,Griffis2007}, or using only the last $N$ years of the data.\footnote{Rich Vogel citation?}
Each model will necessarily give different estimates, which may fit the historical record with comparable accuracy but diverge at an increasing rate as the model is forecast further into the future.
Despite the wide suite of available models, the question of how one should use these nonstationarity estimates for the design of risk mitigation infrastructure or the pricing of insurance instruments has been much less studied.

A critical question in this regard is how to assess whether any structure designed using these projections may actually perform in the future relative to the proposed level of risk protection.
These questions go well beyond what can be reliably answered at this time.
Clearly, in a nonstationary world, the probability of exceedance could be considered to be temporally variable over the project life.
This would however, need to be a prediction of that probability ex ante, using appropriate modeling tools applied to the historical data or to climate model based forecasts.
The risk estimation from the historical record must explicitly inform a risk prediction over the future planning period, and depending on the approach and the form of the underlying nonstationarity, we can expect both bias and uncertainty associated with the prediction, and we need temporally varying probabilities to address the inevitable risk clustering.
This is then the statement of the risk mitigation design problem that we need to consider for nonstationary climate.

In this paper we provide some initial insight into the bias and uncertainty associated with flood frequency estimation for projects as a function of the project planning period $M$ as well as the length of the historical record $N$.
The project planning period can be interpreted as the design life of a physical structure, or as the contract length of a financial instrument.
We focus on structured experiments with synthetic data on extremes to illustrate how one could approach the analyses and what the implications are of varying $N$ and $M$ on the ability to do effective inference and on the ability to judge operational performance.
Though the specific examples target flood risk, this method and framework may be used for understanding the implications of uncertain projections of the future for climate hazards including drought, hurricanes, and wind. \footnote{Overall comment on the introduction:  We may consider hinting at some of the results a bit more, otherwise the introduction is fairly wide open and I'm not sure the reader gets a good sense for what she is in for... but then again, perhaps that's OK}.

@@

To generate $\mu(t)$, a sequence of length $M+N$ is randomly chosen from the \SI{100000}{year} NINO3 sequence.
Next, $\mu(t)$ is calculated deterministically from the NINO3 time series $x(t)$ by
\begin{equation}\label{eq:nino3}
  \mu(t) = \mu_0 + \beta x(t) + \gamma \qty(t - t_0)
\end{equation}
where $\beta$ controls the sensitivity to the NINO3 variation and $\gamma$ controls the magnitude of the secular trend ($\gamma=0$ gives a stationary series).

@@

The variance estimates show similar patterns to those of the bias and are shown in \cref{fig:stationary-variance} as the log standard deviation.
As suggested in the bias plots, the HMM and LN2 Stationary models have very low variance of their estimates.
The LN2 Trend model has much higher variance in estimates, particularly for short $N$ and long $M$.
\begin{figure}[ht]
  \includegraphics[width=\textwidth]{variance_stationary.pdf}
  \caption{The estimation variance for 1000 sequences generated by each of two stationary generating functions (row) and fit by each of three fitting functions (columns) as a function of $M$ and $N$.\label{fig:stationary-variance}}
\end{figure}

@@

\section{Spatial Risk}

Many actors, particularly those in the private sector such as reinsurance vendors, may be exposed to concurrent risk in many different locations.
It is consequently of interest to examine how their risk portfolio risk may vary as a function of time.

We consider that the annual-maximum streamflow in each location depends only on the NINO 3.4 index as shown in \cref{eq:nino3} ($\gamma=0$ to enforce zero secular change).
However, we allow the dependance of streamflow at each location on the NINO 3.4 index to be a random variable, such that $\beta \sim \mathcal{N} \qty(0, 1)$.
This assumption represents observations that during El Ni\~no events drought occurs in some parts of the world and floods in others \citep{Ropelewski1987,Ward2014}.

\Cref{fig:spatial-risk} shows the proportion of locations that experience 100-year floods in each year, empirically defined from a simulated \SI{2000}{year} record at that location.
The orange line shows that if the \gls{iid} assumption is valid, the proportion of sites experiencing a 100-year flood each year is always between 0.005 and 0.015.
However, if the data is simulated from the log-normal distribution conditional on the NINO 3.4 series, as few as zero sites or as many as 20\% of all sites may experience floods in a particular year.
In other words, even though the portfolio has managed its risk by choosing locations with strongly positive, strongly negative, and near-zero dependence on \gls{enso}, the distribution of portfolio risk has a very fat tail, in agreement with previous analyses \citep{Bonnafous2017a}.
\begin{figure}
  \includegraphics[width=\textwidth]{case_study.pdf}
  \caption{
    Proportion of sites experiencing a 100-year flood return in each year.
    (a): A 250-year series of data simulated with the NINO 3.4 model (blue) and using an \gls{iid} model (orange). 
    (b): The probability distribution function of data simulated with the NINO 3.4 model and using an \gls{iid} model.
    This distribution is estimated using a \SI{2000}{year} record, of which the first \SI{250}{year} is shown in (a).\label{fig:spatial-risk}
  }
\end{figure}

@@

Infrastructure in the developed world is aging \citep{Ho2017}, leading to questions of whether and how to replace it.
In the developed world new infrastructure is being built.
Given a positive future discount rate and uncertainty associated with future urbanization, economic growth, and climate change, the expected net benefits of this infrastructure are highly uncertain.
At the same time the private sector has accepted an increasing share of global risk \citep{WorldEconomicForum2016}, which it has mitigated primarily through financial instruments.
How should different contracts be priced? When does it make sense to build large infrastructure?
These questions cannot be answered with existing methodologies.

@@

The approach being taken by many, including the \gls{ipcc}, is to consider discrete scenarios for what humans may do with respect to greenhouse gas emissions, and then attempt to use climate models to estimate the probability of extreme events conditional on occurrence of a particular scenario.
Because of inherent errors due to the discretization and parameterization of the equations of motion, as well as uncertainty about the underlying physical parameters, bias correction methods \citep[\eg quantile-quantile mapping;][]{Rajczak2016} are typically applied to the final outputs.
While this typically improves the model chain's representation of the historical record, this approach assumes a stationary relationship between modeled and observed rainfall at a point-by-point scale, an assumption that is unlikely to hold under the kinds of spatial and dynamical shifts anticipated under increased greenhouse gas scenarios such as expansion of the tropics and altered equator-pole temperature contrasts.\citep{Dankers2009,Ott2013,Merz2014,Dittes2017}.\footnote{Farnham et al 2018!}

The idea that one needs to consider the joint uncertainty in the estimation and occurrence of climate extremes for the $N$ and $M$ periods for stationary conditions was originally considered by \citet{Lall1987}.
That analysis corresponds to what one would expect under \gls{iid} conditions, and would hence provide a baseline to compare the uncertainty distribution associated with a portfolio whose elements were designed using $N$ years of data with an application to $M$ future years at a common design probability of exceedance for the elements.
